Trawl Summary

### 2020-05-13 text/trawl_summary.Rmd
Trawl analysis updated and Common name fixes should be copmlete.

## Beta diversity

We can start to break this down further, and the inference that species richness might be increasing over time has some support (again, not sure yet if this is real or methodological).

@@

Taking this further,

>>>>

## Species diversity

We can start to break this down further, and the inference that species richness might be increasing over time has some support (again, not sure yet if this is real or methodological). 2020 is an obvious outlier here because we have only conducted 3 trawls (bummer) and I don't know what that little blip is in 2014. One weird instance where the code is mischaracterizing something.

@@

## Community changes

Taking this further, we can look at how evenly species are distributed within the community. Rank-abundance plots of the community for integrated across each year show common species are always common (not surprising) and there is a lot of inter-annual variability in the less common species. Here, 'Species rank' is calculated based on the full time series and fixed. We then plot the relative abundance of each species in each year. The black line shows the global community rank-abundance relationship.

```{r rank-abundance plots, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

TB_ann_relN = TB_trawl_data %>% select(year, Common_name, Abundance) %>% na.omit %>%
  group_by(year, Common_name) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE)) %>%
  mutate(rel_N = Abundance/sum(Abundance, na.rm = TRUE)) %>%
  left_join(TB_trawl_data %>% 
              # Turn filter back on to use 2007 as reference year
              # filter(year == 2007) %>%
              na.omit %>%
              group_by(Common_name) %>%
              summarise(Abundance = sum(Abundance, na.rm = TRUE)) %>%
              mutate(rel_N = Abundance/sum(Abundance, na.rm = TRUE)) %>%
              mutate(rank = dense_rank(desc(rel_N))) %>%
              select(Common_name, rank)) %>%
  ungroup() %>%  mutate(year = as.factor(year))

global_relN <- TB_trawl_data %>% 
              # Turn filter back on to use 2007 as reference year
              # filter(year == 2007) %>%
              na.omit %>%
              group_by(Common_name) %>%
              summarise(Abundance = sum(Abundance, na.rm = TRUE)) %>%
              mutate(rel_N = Abundance/sum(Abundance, na.rm = TRUE)) %>%
              mutate(rank = dense_rank(desc(rel_N))) %>%
              select(Common_name, rel_N, rank)

ggplot(TB_ann_relN, aes(x = rank, y = log10(rel_N))) + 
    geom_line(data = global_relN, aes(x = rank, y = log10(rel_N)), size = 2, color = "black") +
  geom_line(aes(group = year, color = year), alpha = 0.7, size = 1.2) +
  scale_y_continuous(name = expression(log[10]*"(Relative Abundance)"))+
  scale_x_continuous(name = "Species Rank")+
  scale_color_viridis(discrete = TRUE) +
  theme_minimal()

These inter-annual trends aside, we see that evenness of the community doesn't show drastic differences among years. Sampling effort is going to explain some of this, but this also might represent some neutral (zero sum) or compensatory dynamics among species in the community.

```{r annual rank-abundance, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

TB_relN_ann = TB_trawl_data %>% select(year, Common_name, Abundance) %>% na.omit %>%
  group_by(year, Common_name) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE)) %>%
  mutate(rel_N = Abundance/sum(Abundance, na.rm = TRUE),
         rank = dense_rank(desc(rel_N))) %>%
  ungroup() %>%  mutate(year = as.factor(year))

ggplot(TB_relN_ann, aes(x = rank, y = log10(rel_N))) + 
  geom_line(aes(group = year, color = year), alpha = 0.6, size = 1.2) +
   scale_y_continuous(name = expression(log[10]*"(Relative Abundance)"))+
  scale_x_continuous(name = "Species Rank")+
  scale_color_viridis(discrete = TRUE) +
  theme_minimal()

We can characterize the how communities shift and the important players (i.e. those with loadings on PC1 or PC2 of the most extreme 1%). This shows there is are noticeable shifts in the community composition over the years and there appears to by directional changes with time.

```{r craig pca, include = TRUE, echo= FALSE, warning = FALSE, message=FALSE}
# create taxa by site matrix with common names
TB_trawl_commonsite <- TB_trawl_data %>%
  select(year, month, date_id, Common_name) %>%
  group_by(date_id, Common_name) %>%
  distinct() %>% 
  mutate(pres = 1) %>%
  pivot_wider(names_from = Common_name, values_from = pres, values_fill = list(pres = 0)) %>%
  ungroup()
##addressing outlier points
TB_trawl_outlier_rm <- TB_trawl_commonsite %>%
  filter(date_id %ni% c("2018-1","2007-7"))

### Legendre PC with Hellinger
# Hellinger pre-transformation of the species matrix
tb.h <- decostand (TB_trawl_outlier_rm %>% select(-c(1:2)) %>% column_to_rownames("date_id"), "hellinger")
tb.h.pca <- rda(tb.h)
tb.h.pca.summ <- summary(tb.h.pca)

#pull coordinates and loadings
TB <- data.frame(matrix(ncol=2,nrow=105, dimnames=list(NULL, c("PC1", "PC2"))))

df1  <- data.frame(tb.h.pca.summ$sites[,1:2])
TB$PC1 <- df1[,1]
TB$PC2 <- df1[,2]
TB$date_id <- row.names(tb.h.pca.summ$sites)
TB <- left_join(TB, TB_trawl_outlier_rm, by=c("date_id"="date_id"))

Var = factor(TB$year, levels =c("2007","2009","2010","2011",
                                "2012","2013","2014","2015",
                                "2016","2017","2018","2019",
                                "2020"))
#pull loadings
TB.loadings  <- data.frame(tb.h.pca.summ$species[,1:2])
TB.loadings$Species <- row.names(tb.h.pca.summ$species)
#filtering out top species contributions
TB.loadings.top <- TB.loadings %>%
  filter(PC1>quantile(PC1, prob=.99) | 
           PC1<quantile(PC1, prob=.01)| 
           PC2>quantile(PC2, prob=.99)  | 
           PC2<quantile(PC2, prob=.01))

#analyses
# tb.h.dl <- dist(tb.h)
# pc_test<- betadisper(tb.h.dl, TB$year)
# anova(pc_test)
# permutest(pc_test, pairwise = TRUE)

TB.centroids <- TB %>%
  group_by(year) %>%
  summarise(
    meanPC1 = mean(PC1),
    meanPC2 = mean(PC2)
  )
TB$year <- as.integer(TB$year)

#figure
ggplot(data = TB, aes(x=PC1, y=PC2, color=as.factor(year))) + 
  geom_point(aes(color=as.factor(year)), size =3, alpha = 0.6)+
  geom_path(data=TB.centroids, aes(x=meanPC1, y=meanPC2), arrow=arrow(), size=1, color="grey20")+
  geom_text(data=TB.centroids, aes(x=meanPC1, y=meanPC2, label=year), color="grey20")+
  ggrepel::geom_text_repel(data=TB.loadings.top, aes(x=PC1, y=PC2, label=Species), color="grey30")+
  geom_hline(yintercept=0, linetype="dotted") +
  geom_vline(xintercept=0, linetype="dotted") +
  labs(color = "Year") +
  theme_minimal() +
  scale_colour_viridis(discrete = TRUE)

The environmental data will need to be completed further to begin to test if these changes are driven by the environment. In its absence, it might be a useful exercise to see how the composition of community traits might guide the expected patterns observed in the environment.

```{r rank table, include = TRUE, echo = FALSE, warning = FALSE, message = FALSE}

rank_table <- TB_relN_ann %>% group_by(Common_name) %>%
  summarise(mean_rank = round(mean(rank),2),
              rank_sd = round(sd(rank, na.rm = TRUE),2)) %>%
  arrange(mean_rank) %>%
  rename(`Common Name` = "Common_name",
         `Mean Rank` = "mean_rank",
         `Rank SD` = "rank_sd")

knitr::kable(rank_table)
